---
title: "ReAnnotateR: Functional Annotation of Genomic Intervals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ReAnnotateR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r load-packages, message=FALSE}
library(ReAnnotateR)

# Check if required packages are available
has_txdb <- requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE)
has_orgdb <- requireNamespace("org.Hs.eg.db", quietly = TRUE)

if (!has_txdb || !has_orgdb) {
  message("Note: Some code chunks require TxDb.Hsapiens.UCSC.hg38.knownGene and org.Hs.eg.db packages.")
  message("Install them with: BiocManager::install(c('TxDb.Hsapiens.UCSC.hg38.knownGene', 'org.Hs.eg.db'))")
}
```

# Introduction

The **ReAnnotateR** package provides tools to annotate genomic intervals from BED files by linking them to genes, genomic features like promoters, exons, introns, intergenic regions, and biological functions. This vignette demonstrates a complete workflow.

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("giills/ReAnnotateR")

# Install required Bioconductor packages
BiocManager::install(c("TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db"))
```

# Load Required Libraries

```{r libraries, message=FALSE, eval=has_txdb && has_orgdb}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
```

# Load Example Data

ReAnnotateR includes example BED files for demonstration. Alternatively, you can load your own BED file.

```{r load-data}
# Load example BED file
bed_file <- system.file("extdata", "example.bed", package = "ReAnnotateR")

# For demonstration, create a simple example if file doesn't exist
if (!file.exists(bed_file)) {
  # Create example data
  library(GenomicRanges)
  gr <- GRanges(
    seqnames = rep("chr1", 5),
    ranges = IRanges::IRanges(
      start = c(10000, 50000, 100000, 150000, 200000),
      end = c(11000, 51000, 101000, 151000, 201000)
    )
  )
} else {
  gr <- read_bed(bed_file)
}

# Display the genomic ranges
print(gr)
```

The `read_bed()` function validates that:
- Start positions are >= 1
- End positions are >= start positions
- The file exists and is properly formatted

# Annotate Genomic Features

The first step is to classify each genomic interval by its feature type: promoter, exon, intron, or intergenic region.

```{r annotate-features, eval=has_txdb}
# Load the TxDb object for hg38
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene::TxDb.Hsapiens.UCSC.hg38.knownGene

# Annotate regions (default: 3kb upstream, 3kb downstream for promoters)
gr_annotated <- annotate_regions(gr, txdb_info = txdb, 
                                  promoter_up = 2000, 
                                  promoter_down = 500)

# View feature distribution
table(mcols(gr_annotated)$feature_type)
```

```{r annotate-features-fallback, eval=!has_txdb, echo=FALSE}
message("TxDb.Hsapiens.UCSC.hg38.knownGene not available - skipping annotation example")
```

You can customize the promoter definition by adjusting `promoter_up` and `promoter_down` parameters.

# Identify Nearest Genes

Next, we can identify the closest gene to each interval and calculate the distance.

```{r nearest-genes, eval=has_txdb}
gr_with_genes <- nearest_gene(gr_annotated, txdb_info = txdb)

# Create a summary table
result_df <- data.frame(
  chr = as.character(seqnames(gr_with_genes)),
  start = start(gr_with_genes),
  end = end(gr_with_genes),
  feature = mcols(gr_with_genes)$feature_type,
  nearest_gene = mcols(gr_with_genes)$nearest_gene,
  distance = mcols(gr_with_genes)$distance_to_gene
)

# Show first few rows
head(result_df)
```

A distance of 0 indicates the interval overlaps the gene body.

# Functional Enrichment Analysis

Perform Gene Ontology (GO) enrichment analysis to identify biological processes associated with the nearest genes.

```{r functional-enrichment, message=FALSE, warning=FALSE, eval=has_txdb && has_orgdb}
# Perform GO enrichment (Biological Process)
enrichment_results <- functional_terms(gr_with_genes, 
                                       orgdb = org.Hs.eg.db::org.Hs.eg.db, 
                                       go = TRUE, 
                                       kegg = FALSE)

# Display top enriched terms
if (!is.null(enrichment_results$go) && nrow(enrichment_results$go) > 0) {
  head(enrichment_results$go[, c("Description", "pvalue", "Count", "GeneRatio")], 10)
} else {
  message("No significant GO terms found (may occur with small datasets)")
}
```

You can also enable KEGG pathway enrichment by setting `kegg = TRUE`!

# Test for Feature Enrichment

Compare your intervals to a background set to test if certain genomic features are enriched using Fisher's exact test.

```{r fisher-enrichment, eval=has_txdb}
# Create a simulated background set for demonstration
# In practice, you should load your actual background regions
set.seed(123)
n_bg <- 1000
bg_chr <- rep("chr1", n_bg)
bg_starts <- sort(sample(1000000:200000000, n_bg))
bg_ends <- bg_starts + sample(500:2000, n_bg, replace = TRUE)

bg_gr <- GenomicRanges::GRanges(
  seqnames = bg_chr,
  ranges = IRanges::IRanges(start = bg_starts, end = bg_ends)
)

bg_gr_annotated <- annotate_regions(bg_gr, txdb_info = txdb)

promoter_enrichment <- fisher_enrichment(gr_annotated, 
                                         category = "promoter", 
                                         background_gr = bg_gr_annotated)

if (!is.null(promoter_enrichment)) {
  cat("Promoter Enrichment Results:\n")
  cat("Odds Ratio:", round(promoter_enrichment$odds_ratio, 3), "\n")
  cat("P-value:", format.pval(promoter_enrichment$p_value), "\n\n")
  print(promoter_enrichment$table)
}
```

An odds ratio > 1 indicates enrichment, while < 1 indicates depletion.

# Coordinate Conversion (Optional)

If you need to convert coordinates between genome assemblies (e.g., hg19 to hg38):

```{r convert-coords, eval=FALSE}
# Example: Convert from hg19 to hg38
chain_file <- "path/to/hg19ToHg38.over.chain"
gr_hg38 <- convert_coordinates(gr, chain_file)
```

# Export Results

Save your annotated results for downstream analysis.

```{r export, eval=FALSE}
# Export annotated GRanges object
export_results(gr_with_genes, "annotated_regions.tsv", format = "tsv")

# Export as BED file
export_results(gr_with_genes, "annotated_regions.bed", format = "bed")

# Export enrichment results
if (!is.null(enrichment_results)) {
  export_results(enrichment_results, "enrichment_results.tsv", format = "tsv")
}
```

# Visualization

ReAnnotateR includes several visualization functions to explore your data:

```{r viz-setup, eval=has_txdb, message=FALSE}
library(ggplot2)
```

## Feature Composition

```{r plot-features, eval=has_txdb, fig.width=7, fig.height=5}
plot_feature_composition(gr_annotated)
```

## Chromosomal Density

```{r plot-density, eval=has_txdb, fig.width=7, fig.height=5}
plot_chromosomal_density(gr_annotated, bin_size = 1e6)
```

## Annotation Summary

```{r plot-summary, eval=has_txdb, fig.width=7, fig.height=5}
plot_annotation_summary(gr_annotated)
```

# Session Info

```{r session-info}
sessionInfo()
```