---
title: "ReAnnotateR: Functional Annotation of Genomic Intervals"
author: "Julia Gilley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ReAnnotateR: Functional Annotation of Genomic Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r load-packages, message=FALSE}
library(ReAnnotateR)

# Check if required packages are available
has_txdb <- requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE)
has_orgdb <- requireNamespace("org.Hs.eg.db", quietly = TRUE)

if (!has_txdb || !has_orgdb) {
  message("Note: Some code chunks require TxDb.Hsapiens.UCSC.hg38.knownGene and org.Hs.eg.db packages.")
  message("Install them with: BiocManager::install(c('TxDb.Hsapiens.UCSC.hg38.knownGene', 'org.Hs.eg.db'))")
}
```

# Introduction

The **ReAnnotateR** package provides tools to annotate genomic intervals from BED files by linking them to genes, genomic features like promoters, exons, introns, intergenic regions, and biological functions. This vignette demonstrates a complete workflow.

This package builds upon established Bioconductor infrastructure including GenomicRanges (Lawrence et al., 2013) for genomic interval manipulation, and clusterProfiler (Wu et al., 2021) for functional enrichment analysis.

## References

- Lawrence M, Huber W, Pagès H, Aboyoun P, Carlson M, et al. (2013) Software for Computing and Annotating Genomic Ranges. *PLoS Computational Biology* 9(8): e1003118. https://doi.org/10.1371/journal.pcbi.1003118

- Wu T, Hu E, Xu S, Chen M, Guo P, Dai Z, Feng T, Zhou L, Tang W, Zhan L, Fu X, Liu S, Bo X, Yu G (2021). "clusterProfiler 4.0: A universal enrichment tool for interpreting omics data." *The Innovation*, 2(3), 100141. https://doi.org/10.1016/j.xinn.2021.100141

- Wickham H (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York. ISBN 978-3-319-24277-4. https://ggplot2.tidyverse.org

- Kent WJ, Sugnet CW, Furey TS, Roskin KM, Pringle TH, Zahler AM, Haussler D (2002). "The human genome browser at UCSC." *Genome Research*, 12(6):996-1006. https://doi.org/10.1101/gr.229102

- Fisher RA (1922). "On the interpretation of χ² from contingency tables, and the calculation of P". *Journal of the Royal Statistical Society*, 85(1): 87-94. https://doi.org/10.2307/2340521

- Carlson M (2019). *TxDb.Hsapiens.UCSC.hg38.knownGene: Annotation package for TxDb object(s)*. R package version 3.10.0.

- Carlson M (2019). *org.Hs.eg.db: Genome wide annotation for Human*. R package version 3.8.2.

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("giills/ReAnnotateR", build_vignettes = TRUE)

# Install required Bioconductor packages
BiocManager::install(c("TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db"))
```

# Load Required Libraries

```{r libraries, eval=has_txdb && has_orgdb}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
```

# Load Example Data

ReAnnotateR includes example BED files for demonstration. Alternatively, you can load your own BED file.

```{r load-data}
# Load example BED file
bed_file <- system.file("extdata", "example.bed", package = "ReAnnotateR")

# For demonstration, create a simple example if file doesn't exist
if (!file.exists(bed_file)) {
  # Create example data
  library(GenomicRanges)
  gr <- GRanges(
    seqnames = rep("chr1", 5),
    ranges = IRanges::IRanges(
      start = c(10000, 50000, 100000, 150000, 200000),
      end = c(11000, 51000, 101000, 151000, 201000)
    )
  )
} else {
  gr <- read_bed(bed_file)
}

# Display the genomic ranges
print(gr)
```

The `read_bed()` function validates that:
- Start positions are >= 1
- End positions are >= start positions
- The file exists and is properly formatted

# Annotate Genomic Features

The first step is to classify each genomic interval by its feature type: promoter, exon, intron, or intergenic region.

```{r annotate-features, eval=has_txdb}
# Load the TxDb object for hg38
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene::TxDb.Hsapiens.UCSC.hg38.knownGene

# Annotate regions (default: 3kb upstream, 3kb downstream for promoters)
# Note: Warnings about out-of-bound ranges on alternative contigs are expected
# and can be safely ignored for standard chromosomes
suppressWarnings({
  gr_annotated <- annotate_regions(gr, txdb_info = txdb, 
                                    promoter_up = 2000, 
                                    promoter_down = 500)
})

# View feature distribution
table(mcols(gr_annotated)$feature_type)
```

```{r annotate-features-fallback, eval=!has_txdb, echo=FALSE}
message("TxDb.Hsapiens.UCSC.hg38.knownGene not available - skipping annotation example")
```

You can customize the promoter definition by adjusting `promoter_up` and `promoter_down` parameters.

# Identify Nearest Genes

Next, we can identify the closest gene to each interval and calculate the distance.

```{r nearest-genes, eval=has_txdb}
gr_with_genes <- nearest_gene(gr_annotated, txdb_info = txdb)

# Create a summary table
result_df <- data.frame(
  chr = as.character(seqnames(gr_with_genes)),
  start = start(gr_with_genes),
  end = end(gr_with_genes),
  feature = mcols(gr_with_genes)$feature_type,
  nearest_gene = mcols(gr_with_genes)$nearest_gene,
  distance = mcols(gr_with_genes)$distance_to_gene
)

# Show first few rows
head(result_df)
```

A distance of 0 indicates the interval overlaps the gene body.

# Functional Enrichment Analysis

Perform Gene Ontology (GO) enrichment analysis to identify biological processes associated with the nearest genes.

```{r functional-enrichment, eval=has_txdb && has_orgdb}
# Perform GO enrichment (Biological Process)
enrichment_results <- functional_terms(gr_with_genes, 
                                       orgdb = org.Hs.eg.db::org.Hs.eg.db, 
                                       go = TRUE, 
                                       kegg = FALSE)

# Display top enriched terms
if (!is.null(enrichment_results$go) && nrow(enrichment_results$go) > 0) {
  head(enrichment_results$go[, c("Description", "pvalue", "Count", "GeneRatio")], 10)
} else {
  message("No significant GO terms found (may occur with small datasets)")
}
```

You can also enable KEGG pathway enrichment by setting `kegg = TRUE`!

# Test for Feature Enrichment

Compare your intervals to a background set to test if certain genomic features are enriched using Fisher's exact test (Fisher, 1922).

```{r fisher-enrichment, eval=has_txdb}
# Create a simulated background set for demonstration
# In practice, you should load your actual background regions
set.seed(123)
n_bg <- 1000
bg_chr <- rep("chr1", n_bg)
bg_starts <- sort(sample(1000000:200000000, n_bg))
bg_ends <- bg_starts + sample(500:2000, n_bg, replace = TRUE)

bg_gr <- GenomicRanges::GRanges(
  seqnames = bg_chr,
  ranges = IRanges::IRanges(start = bg_starts, end = bg_ends)
)

suppressWarnings({
  bg_gr_annotated <- annotate_regions(bg_gr, txdb_info = txdb)
})

promoter_enrichment <- fisher_enrichment(gr_annotated, 
                                         category = "promoter", 
                                         background_gr = bg_gr_annotated)

if (!is.null(promoter_enrichment)) {
  cat("Promoter Enrichment Results:\n")
  cat("Odds Ratio:", round(promoter_enrichment$odds_ratio, 3), "\n")
  cat("P-value:", format.pval(promoter_enrichment$p_value), "\n\n")
  print(promoter_enrichment$table)
}
```

An odds ratio > 1 indicates enrichment, while < 1 indicates depletion.

# Coordinate Conversion (Optional)

If you need to convert coordinates between genome assemblies (e.g., hg19 to hg38), the package uses UCSC liftOver chain files (Kent et al., 2002):

```{r convert-coords, eval=FALSE}
# Example: Convert from hg19 to hg38
chain_file <- "path/to/hg19ToHg38.over.chain"
gr_hg38 <- convert_coordinates(gr, chain_file)
```

# Export Results

Save your annotated results for downstream analysis.

```{r export, eval=FALSE}
# Export annotated GRanges object
export_results(gr_with_genes, "annotated_regions.tsv", format = "tsv")

# Export as BED file
export_results(gr_with_genes, "annotated_regions.bed", format = "bed")

# Export enrichment results
if (!is.null(enrichment_results)) {
  export_results(enrichment_results, "enrichment_results.tsv", format = "tsv")
}
```

# Visualization

ReAnnotateR includes several visualization functions based on ggplot2 (Wickham, 2016) to explore your data. All visualization functions return ggplot2 objects that can be further customized, saved with `ggsave()`, or displayed directly.

```{r viz-setup, eval=has_txdb}
library(ggplot2)
```

## Feature Composition

```{r plot-features, eval=has_txdb, fig.width=7, fig.height=5}
p1 <- plot_feature_composition(gr_annotated)
print(p1)

# You can customize the returned plot
p1 + labs(title = "My Custom Feature Plot") + 
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Chromosomal Density

This visualization shows the distribution of genomic intervals across chromosomes.

```{r plot-density, eval=has_txdb, fig.width=10, fig.height=8}
# Create the density plot with 10 Mb bins
# This shows how intervals cluster along each chromosome
p2 <- plot_chromosomal_density(gr_annotated, bin_size = 10e6)
print(p2)

# You can adjust bin size to see different levels of detail
# Smaller bins (e.g., 5 Mb) show more fine-grained patterns:
# p2_small <- plot_chromosomal_density(gr_annotated, bin_size = 5e6)
# Larger bins (e.g., 20 Mb) show broader trends:
# p2_large <- plot_chromosomal_density(gr_annotated, bin_size = 20e6)
```

## Enrichment Visualization

```{r plot-enrichment, eval=has_txdb && has_orgdb, fig.width=7, fig.height=5}
# Plot enrichment results if available
if (!is.null(enrichment_results$go) && nrow(enrichment_results$go) > 0) {
  p3 <- plot_enrichment(enrichment_results$go, top_n = 10)
  print(p3)
} else {
  message("No enrichment results to plot for this example dataset")
  
  # Show what the plot would look like with simulated data
  simulated_enrichment <- data.frame(
    Description = c("DNA repair", "cell cycle", "apoptosis", 
                    "signal transduction", "transcription"),
    p.adjust = c(0.001, 0.005, 0.01, 0.02, 0.04)
  )
  p3 <- plot_enrichment(simulated_enrichment, top_n = 5)
  print(p3)
}
```

## Annotation Summary

```{r plot-summary, eval=has_txdb, fig.width=7, fig.height=5}
p4 <- plot_annotation_summary(gr_annotated)
print(p4)
```

## Saving Plots

All plots can be saved using `ggsave()`:

```{r save-plots, eval=FALSE}
# Save a single plot
ggsave("feature_composition.png", plot = p1, width = 7, height = 5)

# Save all plots
ggsave("chromosomal_density.png", plot = p2, width = 10, height = 6)
ggsave("enrichment.png", plot = p3, width = 8, height = 6)
ggsave("annotation_summary.png", plot = p4, width = 7, height = 5)
```

# Session Info

```{r session-info}
sessionInfo()
```